Creating figures for the paper

### 2024-07-08 infercnv/reports/paper_figures.Rmd
add schema, update to use qc'ed cells only, add more figures

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# module load R/4.4.0 ; Rscript -e "rmarkdown::render('reports/paper_figures.Rmd', output_dir = 'out/paper_figures/')" # nolint: line_length_linter.

@@

First, we load the CITE-seq object. We convert it to a SingleCellExperiment 
object. Then we add the CNV metadata from the inferCNV run to the object.

```{r load_sce, results = 'asis', class.source = 'fold-show'}
# load sce with gains
sce <-
  "out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/" %>%
  list.files(pattern = "sce_w_gains", full.names = TRUE) %>%
  readRDS()
```

# Plot distribution of number of fibroblasts per patient to decide cut-off

```{r plot_fibroblast_dist}
n_fibroblasts <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::count(sample)
n_fibroblasts %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 20)
```

I will introduce a cut-off of 100 fibroblasts per patient.

```{r cut_off}
n_fibroblasts %>%
  dplyr::summarise(`n >= 100` = sum(n >= 100),
                   `n < 100` = sum(n < 100),
                   n = dplyr::n()) %>%
  knitr::kable()
```

This leaves 52 / 80 of the samples.

# Plot distribution of chr7+ and chr5+ rates

```{r}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, chr7_gain, chr5_gain) %>%
  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(event = dplyr::case_when(all(is.na(value) | value == 0) ~ "WT",
                                         value == 1 ~ name, TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ","),
    n_events = sum(value, na.rm = TRUE)) %>%
  # get props and counts
  dplyr::ungroup() %>%
  dplyr::add_count(sample, name = "n_cells_per_sample") %>%
  dplyr::add_count(sample, events, name = "n_cells_per_sample_x_event") %>%
  dplyr::group_by(sample, n_cells_per_sample_x_event, n_cells_per_sample, events) %>%
  dplyr::summarise(prop_cells_per_sample_x_event = n_cells_per_sample_x_event / n_cells_per_sample) %>%
  dplyr::distinct() %>%
  dplyr::mutate(sample = forcats::fct_reorder(sample, prop_cells_per_sample_x_event)) %>%
  tidyr::pivot_longer(c("n_cells_per_sample_x_event", "prop_cells_per_sample_x_event"))

# get ordering by total non-WT prop

pdf("test.pdf")
events %>%
  ggplot(aes(x = sample, y = value, fill = events)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  facet_wrap(~ name, scale = "free_y", ncol = 1)
dev.off()
```

@@

```{r ct_comp, fig.width = 12, fig.height = 7}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    sample = forcats::fct_infreq(sample),
    cell_group = forcats::fct_rev(forcats::fct_infreq(cell_group))) %>%
  dplyr::add_count(sample, cell_group, name = 'n_sample_x_cell_group') %>%
  dplyr::add_count(cell_group, name = 'n_cell_group')
p_base <-
  p_dat %>%
  ggplot(aes(fill = cell_group, colour = cell_group)) +
  scale_fill_manual(values = dittoSeq::dittoColors()) +
  scale_colour_manual(values = dittoSeq::dittoColors())
p1 <- 
  p_base +
  geom_point(aes(x = sample, y = cell_group, 
                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
  theme_classic() +
  theme(legend.position = 'none',
        axis.text.x = element_text(size = 7)) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  labs(y = params$cell_group_name)
p2 <- 
  p_base +
  geom_bar(aes(x = sample), position = 'stack') +
  theme_void() +
  theme(legend.position = 'none',
        axis.text.y = element_text()) 
p3 <-
  p_base +
  geom_bar(aes(y = cell_group), position = 'stack') +
  geom_text(
    data = p_dat %>% dplyr::distinct(cell_group, n_cell_group),
    aes(x = n_cell_group, y = cell_group, 
        label = format(n_cell_group, nsmall = 1, big.mark = ",")), 
    hjust = -0.1) +
  theme_void() +
  theme(legend.position = 'none') +
  xlim(0, 1.2 * max(p_dat$n_cell_group))
design <-
'BB#
BB#
AAC'

@@

pal <-
  c('T cell' = "#E6AB02",
    'Stromal cell' = "#66A61E",
    'Myeloid cell' = "#E7298A",
    'B cell/plasma cell' = "#7570B3",
    'Endothelial cell' = "#D95F02",
    'NK' = "#1B9E77")
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
  coord_fixed() +
  scale_colour_manual(values = pal)
```

@@

```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = 'hide'}
paste0('chr', 1:22) %>% 
  purrr::map(function(chr) {
    events <-
      paste('proportion', c('dupli', 'loss'), chr, sep = '_')
    reducedDim(sce, 'UMAP_uwot') %>%
      cbind(sce@colData[, events]) %>%
      tibble::as_tibble() %>%
      tidyr::pivot_longer(3:4) %>%
      dplyr::arrange(value) %>%
      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
      geom_point(size = 0.1) +
      coord_fixed() +
      theme_classic() +
      scale_color_viridis_c() +
      facet_grid(~ name) %>%
      print()
  })
```

@@

```{r plot_gains_and_losses_by_ct, results = 'hide'}
# at single-cell level
sce@colData[, params$cell_group_name] %>%
  unique() %>%
  purrr::map(function(cgn) {
    c('dupli', 'loss') %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[sce@colData[, params$cell_group_name] == cgn, ]
        if (nrow(p_dat) > 0) {
          p_mat <-
            p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
          p_annot <- 
            p_dat[, c('sample'), drop = F] %>%
            as.data.frame()
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ((nrow(p_mat) < 65536) == T),
            legend = F,
            main = paste0(params$cell_group_name, ': ', cgn, '\ndirection: ', direction)) %>%
            print()
        }
      })
  })
```

@@

```{r plot_gains_and_losses_by_id, fig.width = 12}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(cell_group = get(params$cell_group_name)) %>%
  dplyr::group_by(sample, cell_group) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(sample_x_cg = paste0(sample, '_', cell_group)) %>%
  tibble::column_to_rownames('sample_x_cg')

# split gain and loss
c('dupli', 'loss') %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c('sample', 'cell_group')] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0('direction: ', direction)) %>%
      print()
  })
```

@@

  # plot both count (stack) and proportion (fill), annotate n cells
  list(
    plot_func(p_dat, position_fill(reverse = T)) +
      ggtitle(title) +
      theme(axis.text.x = element_blank(),
            axis.title.x = element_blank()) +
      ylab('count'),
    plot_func(p_dat, position_stack(reverse = T)) +
      ylab('proportion')) %>%
    patchwork::wrap_plots(ncol = 1) 
}

# plot, across all and by patient
plot_prop_chr7_gain(p_dat, 'chr7 gain - all cells')
p_dat %>%
  {split(., .$sample)} %>%
  purrr::walk2(names(.), ., function(id, df) {
    df %>%
    plot_prop_chr7_gain(
      paste0('chr7 gains - ', id, ' (', unique(df$n_cells), ' cells)')) %>%
      print()
  })
```

@@

```{r assign_gains, class.source = 'fold-show'}
sce <- xfun::cache_rds({

  # apply cut-offs for whole chr gains and losses
  paste0('chr', 1:22) %>%
    purrr::walk(function(chr) {
      chr_dupli <-
        sce@colData[, paste0("proportion_dupli_", chr)]
      sce@colData[, paste0(chr, '_gain')] <<-
        ifelse(chr_dupli > params$min_prop_for_cnv, 1, 
               ifelse(chr_dupli < params$max_prop_for_wt, 0, 
                      NA))
      chr_loss <-
        sce@colData[, paste0("proportion_loss_", chr)]
      sce@colData[, paste0(chr, '_loss')] <<-
        ifelse(chr_loss > params$min_prop_for_cnv, 1, 
               ifelse(chr_loss < params$max_prop_for_wt, 0, 
                      NA))
    })

@@

}, file = 'sce_w_gains.rds', rerun = params$rerun)
```

```{r plot_heatmap, fig.width = 10, fig.height = 5}
event_cols <-
  c('WT' = 'grey',
    'chr5_gain' = '#e7298a',
    'chr5_gain,chr7_gain' = '#d95f02',
    'chr7_gain' = '#e6ab02')

p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, 
                dplyr::ends_with('_gain'), 
                dplyr::ends_with('_loss')) %>%
  tidyr::pivot_longer(-c('sample', 'cell')) %>%
  dplyr::filter(name %in% c('chr7_gain', 'chr5_gain')) %>%
  dplyr::group_by(sample, cell) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ 'WT',
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ',')) %>%
  dplyr::mutate(events = forcats::fct_relevel(events, names(event_cols))) %>%
  # get n aneuploid for plot ordering
  dplyr::group_by(sample) %>%
  dplyr::mutate(prop_aneuploid = sum(events != 'WT') / dplyr::n()) %>%
  # count
  dplyr::group_by(sample, prop_aneuploid) %>%
  dplyr::count(events) %>%
  # for fill and stack
  tidyr::crossing(tibble::tibble(plot = c('% cells', 'n cells')))

# plot
p_dat %>%
  ggplot(aes(x = reorder(sample, -prop_aneuploid), y = n, fill = events)) +
  geom_col(data = p_dat %>% dplyr::filter(plot == '% cells'),
           position = 'fill') +
  geom_col(data = p_dat %>% dplyr::filter(plot == 'n cells'),
           position = 'stack') +
  theme_classic() +
  ggh4x::facet_grid2(plot ~ ., scales = 'free_y', independent = 'y') +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = event_cols) +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(size = 6))
```

@@

```{r chr7_and_others}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with('_gain'), 
                dplyr::ends_with('_loss')) %>%
  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ 'WT',
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ','),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <= 25 cells
  dplyr::filter(events != 'WT', n >= 25) %>%
  dplyr::mutate(
    events = events %>% 
      gsub('chr', '', .) %>% 
      gsub('_gain', '+', .) %>%
      gsub('_loss', '-', .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
    ))

# plot
p <-
  p_dat %>%
  ggplot(aes(x = events)) +
  geom_text(aes(y = n,  label = label), 
            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'bottom')
p + 
  geom_bar(aes(fill = sample))
p + 
  geom_bar(aes(fill = cell_group)) +
  theme(legend.position = 'none')
```

@@

```{r cont_tab}
cont <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, 'chr7+', 'chr7_WT'),
                chr5 = ifelse(chr5_gain == 1, 'chr5+', 'chr5_WT')) 
table(cont$chr7, cont$chr5) %>%
  knitr::kable()
```

@@

  fisher <-
    colnames(sce@colData)[grepl('_gain$|_loss$', colnames(sce@colData))] %>%
    purrr::set_names() %>%
    purrr::map(function(event) {

@@

        # get contingencies
        cont <-
          table(sce@colData[, c('chr7_gain', event)])

         # return relevant stats
        tibble::tibble(
          OR = fisher$estimate, 
          p = fisher$p.value,
          sig = ifelse(fisher$p.value < 0.05, '*', ''),
          n_WT = cont[1, 1],
          n_chr7_gain = cont[2, 1],
          n_event = cont[1, 2],
          n_both = cont[2, 2])

@@

    }) %>%
    dplyr::bind_rows(.id = 'event') %>%
    # remove chr7+ (can't do chr7+ vs chr7+)
    dplyr::filter(event != 'chr7_gain', n_event > 0) %>%
    dplyr::arrange(p)

@@

}, file = 'fisher.rds', rerun = T)

@@

# create glm
# prop_chr7_gain ~ age (+ coverage) (per donor)
for_glm <-
  dplyr::inner_join(
    tibble::as_tibble(sce@colData), 
    metadata, 
    by = c('sample' = 'id')) %>%
  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
  dplyr::group_by(sample, age, CTAP, sex, race, ethnicity) %>%
  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
                   n_cells = dplyr::n(),
                   ch7_gain_prop = n_chr7_gain / n_cells)

@@

# all categorical variables
for_glm %>%
  tidyr::pivot_longer(cols = c('CTAP', 'sex', 'race', 'ethnicity')) %>%
  dplyr::group_by(name, value) %>%
  dplyr::mutate(med = median(ch7_gain_prop)) %>%
  ggplot(aes(
    x = tidytext::reorder_within(value, by = med, within = name), 
    y = ch7_gain_prop)) +
  geom_boxplot() +
  geom_jitter(height = 0) +
  facet_wrap(~ name, nrow = 1, scales = 'free') +
  tidytext::scale_x_reordered(guide = guide_axis(angle = -45)) +
  theme_classic() +
  labs(x = '')

>>>>

```{r setup, include = F, message = F, warning = F, class.source = "fold-hide"}
# module load R/4.4.0 ; Rscript -e "rmarkdown::render("reports/paper_figures.Rmd", output_dir = "out/paper_figures/")" # nolint: line_length_linter.

@@

First, we load the CITE-seq data as a `SingleCellExperiment` object. The CNV 
metadata from the `inferCNV` run has been aded to this object.

```{r load_sce, results = "asis", class.source = "fold-show"}
# load sce with gains
sce <-
  "out/Zhang2023/by_fibroblast_cluster/summary/gains_and_losses_cache/" %>%
  list.files(pattern = "sce_w_gains", full.names = TRUE) %>%
  readRDS()
```

# Quantifying CNVs of interest in the samples' fibroblast populations

We first prepare the data so that we can view the proportions of chr7 and chr5 
gains in each patient's fibroblast population.

```{r}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, chr7_gain, chr5_gain) %>%
  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(event = dplyr::case_when(all(is.na(value) | value == 0) ~ "WT",
                                         value == 1 ~ name, TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ",") %>%
      factor(levels = c("chr5_gain,chr7_gain", "chr7_gain", "chr5_gain", "WT")),
    n_events = sum(value, na.rm = TRUE)) %>%
  # get props and counts
  dplyr::ungroup() %>%
  dplyr::add_count(sample, name = "n_cells_per_sample") %>%
  dplyr::add_count(sample, events, name = "n_cells_per_sample_x_event") %>%
  dplyr::distinct(sample, events,
                  n_cells_per_sample_x_event, n_cells_per_sample) %>%
  dplyr::mutate(prop_cells_per_sample_x_event = n_cells_per_sample_x_event /
                  n_cells_per_sample) %>%
  # refactor samples by prop non-WT cells
  dplyr::group_by(sample) %>%
  dplyr::mutate(
    n_wt_per_sample = dplyr::case_when(events == "WT" ~
                                         n_cells_per_sample_x_event,
                                       TRUE ~ 0),
    prop_non_wt_per_sample = (n_cells_per_sample - sum(n_wt_per_sample)) /
                  n_cells_per_sample) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(sample = forcats::fct_reorder2(sample, n_cells_per_sample,
                                               prop_non_wt_per_sample)) %>%
  tidyr::pivot_longer(c("n_cells_per_sample_x_event",
                        "prop_cells_per_sample_x_event"))
```

# Filtering out patients with few fibroblasts

```{r}
min_n_fibroblasts_per_sample <- 100
```

We set the minimum number of fibroblasts per patient to 
`r min_n_fibroblasts_per_sample`. We can review how many patients pass the 
cut-off to give us reliable results.

```{r plot_fibroblast_dist}
n_fibroblasts <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::count(sample)
n_fibroblasts %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 20) +
  geom_vline(xintercept = min_n_fibroblasts_per_sample, linetype = "dashed", 
             colour = "red") +
  theme_classic()
```

```{r cut_off}
enough_fibroblasts <-
  dplyr::filter(n_fibroblasts, n >= min_n_fibroblasts_per_sample)$sample
n_fibroblasts %>%
  dplyr::summarise(`n >= 100` = sum(n >= 100),
                   `n < 100` = sum(n < 100),
                   n = dplyr::n()) %>%
  knitr::kable()
```

This leaves `r length(enough_fibroblasts)` / `r nrow(n_fibroblasts)` of the 
samples. We will filter the `SingleCellExperiment` object accordingly.

We can also review the relationship between the number of cells in each 
patient's fibroblast population and the proportions / counts of the CNVs of 
interest (chr7 gain and chr5 gain).

```{r plot_event_vs_fibroblast}
events %>%
  ggplot(aes(x = sample, y = value, fill = events)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  facet_wrap(~ name, scale = "free_y", ncol = 1) +
  theme_classic() +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")
```

We can now apply the cut-off.

```{r apply_cutoff}
sce <- sce[, sce$sample %in% enough_fibroblasts]
events <- events %>% dplyr::filter(sample %in% enough_fibroblasts)
```

After applying the cut-off, we look for sample-size effects once again.

```{r plot_event_vs_fibroblast_after_cutoff}
events %>%
  ggplot(aes(x = sample, y = value, fill = events)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  facet_wrap(~ name, scale = "free_y", ncol = 1) +
  theme_classic() +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")
```

We can see that there is still a strong relationship between the size of the
fibroblast populations by patient and the proportion of those cells that are
called as having a chr7 or chr5 gain. This is most likely related to with the
fact that we set `ref_group_names = NULL`. This compounds the effect of the
number of cells on the ability to detect CNVs (the false negative rate), 
because...

1. fewer cells makes it less likely to capture these somewhat rare events

2. fewer cells makes the reference profile noisier and less representative of 
the 'baseline' expression levels across chromosomes, further limiting CNV 
calling

Keeping this in mind, we will proceed.

@@

```{r ct_comp, fig.width = 12, fig.height = 7}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(
    sample = forcats::fct_infreq(sample),
    cell_group = forcats::fct_rev(forcats::fct_infreq(cell_group))) %>%
  dplyr::add_count(sample, cell_group, name = "n_sample_x_cell_group") %>%
  dplyr::add_count(cell_group, name = "n_cell_group")
p_base <-
  p_dat %>%
  ggplot(aes(fill = cell_group, colour = cell_group)) +
  scale_fill_manual(values = dittoSeq::dittoColors()) +
  scale_colour_manual(values = dittoSeq::dittoColors())
p1 <- 
  p_base +
  geom_point(aes(x = sample, y = cell_group,
                 alpha = n_sample_x_cell_group, size = n_sample_x_cell_group)) +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 7)) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  labs(y = params$cell_group_name)
p2 <- 
  p_base +
  geom_bar(aes(x = sample), position = "stack") +
  theme_void() +
  theme(legend.position = "none",
        axis.text.y = element_text()) 
p3 <-
  p_base +
  geom_bar(aes(y = cell_group), position = "stack") +
  geom_text(
    data = p_dat %>% dplyr::distinct(cell_group, n_cell_group),
    aes(x = n_cell_group, y = cell_group, 
        label = format(n_cell_group, nsmall = 1, big.mark = ",")), 
    hjust = -0.1) +
  theme_void() +
  theme(legend.position = "none") +
  xlim(0, 1.2 * max(p_dat$n_cell_group))
design <-
"BB#
BB#
AAC"

@@

pal <-
  c("T cell" = "#E6AB02",
    "Stromal cell" = "#66A61E",
    "Myeloid cell" = "#E7298A",
    "B cell/plasma cell" = "#7570B3",
    "Endothelial cell" = "#D95F02",
    "NK" = "#1B9E77")
dittoSeq::dittoDimPlot(sce, "cell_type", size = 0.2) +
  coord_fixed() +
  scale_colour_manual(values = pal)
```

@@

```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = "hide"}
paste0("chr", 1:22) %>% 
  purrr::map(function(chr) {
    events <-
      paste("proportion", c("dupli", "loss"), chr, sep = "_")
    reducedDim(sce, "UMAP_uwot") %>%
      cbind(sce@colData[, events]) %>%
      tibble::as_tibble() %>%
      tidyr::pivot_longer(3:4) %>%
      dplyr::arrange(value) %>%
      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
      geom_point(size = 0.1) +
      coord_fixed() +
      theme_classic() +
      scale_color_viridis_c() +
      facet_grid(~ name) %>%
      print()
  })
```

@@

```{r plot_gains_and_losses_by_ct, results = "hide"}
# at single-cell level
sce@colData[, params$cell_group_name] %>%
  unique() %>%
  purrr::map(function(cgn) {
    c("dupli", "loss") %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[sce@colData[, params$cell_group_name] == cgn, ]
        if (nrow(p_dat) > 0) {
          p_mat <-
            p_dat[, grepl(paste0("proportion_", direction), colnames(p_dat))]
          p_annot <- 
            p_dat[, c("sample"), drop = F] %>%
            as.data.frame()
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ((nrow(p_mat) < 65536) == T),
            legend = F,
            main = paste0(params$cell_group_name, ": ", cgn, "\ndirection: ", direction)) %>%
            print()
        }
      })
  })
```

@@

```{r plot_gains_and_losses_by_id, fig.width = 12}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(cell_group = get(params$cell_group_name)) %>%
  dplyr::group_by(sample, cell_group) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with("proportion_"), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(sample_x_cg = paste0(sample, "_", cell_group)) %>%
  tibble::column_to_rownames("sample_x_cg")

# split gain and loss
c("dupli", "loss") %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0("proportion_", direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c("sample", "cell_group")] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0("direction: ", direction)) %>%
      print()
  })
```

@@

  # plot both count (stack) and proportion (fill), annotate n cells
  list(
    plot_func(p_dat, position_fill(reverse = T)) +
      ggtitle(title) +
      theme(axis.text.x = element_blank(),
            axis.title.x = element_blank()) +
      ylab("count"),
    plot_func(p_dat, position_stack(reverse = T)) +
      ylab("proportion")) %>%
    patchwork::wrap_plots(ncol = 1) 
}

# plot, across all and by patient
plot_prop_chr7_gain(p_dat, "chr7 gain - all cells")
p_dat %>%
  {split(., .$sample)} %>%
  purrr::walk2(names(.), ., function(id, df) {
    df %>%
    plot_prop_chr7_gain(
      paste0("chr7 gains - ", id, " (", unique(df$n_cells), " cells)")) %>%
      print()
  })
```

@@

```{r assign_gains, class.source = "fold-show"}
sce <- xfun::cache_rds({

  # apply cut-offs for whole chr gains and losses
  paste0("chr", 1:22) %>%
    purrr::walk(function(chr) {
      chr_dupli <-
        sce@colData[, paste0("proportion_dupli_", chr)]
      sce@colData[, paste0(chr, "_gain")] <<-
        ifelse(chr_dupli > params$min_prop_for_cnv, 1, 
               ifelse(chr_dupli < params$max_prop_for_wt, 0, 
                      NA))
      chr_loss <-
        sce@colData[, paste0("proportion_loss_", chr)]
      sce@colData[, paste0(chr, "_loss")] <<-
        ifelse(chr_loss > params$min_prop_for_cnv, 1, 
               ifelse(chr_loss < params$max_prop_for_wt, 0, 
                      NA))
    })

@@

}, file = "sce_w_gains.rds", rerun = params$rerun)
```

```{r plot_heatmap, fig.width = 10, fig.height = 5}
event_cols <-
  c("WT" = "grey",
    "chr5_gain" = "#e7298a",
    "chr5_gain,chr7_gain" = "#d95f02",
    "chr7_gain" = "#e6ab02")

p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, 
                dplyr::ends_with("_gain"), 
                dplyr::ends_with("_loss")) %>%
  tidyr::pivot_longer(-c("sample", "cell")) %>%
  dplyr::filter(name %in% c("chr7_gain", "chr5_gain")) %>%
  dplyr::group_by(sample, cell) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ "WT",
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ",")) %>%
  dplyr::mutate(events = forcats::fct_relevel(events, names(event_cols))) %>%
  # get n aneuploid for plot ordering
  dplyr::group_by(sample) %>%
  dplyr::mutate(prop_aneuploid = sum(events != "WT") / dplyr::n()) %>%
  # count
  dplyr::group_by(sample, prop_aneuploid) %>%
  dplyr::count(events) %>%
  # for fill and stack
  tidyr::crossing(tibble::tibble(plot = c("% cells", "n cells")))

# plot
p_dat %>%
  ggplot(aes(x = reorder(sample, -prop_aneuploid), y = n, fill = events)) +
  geom_col(data = p_dat %>% dplyr::filter(plot == "% cells"),
           position = "fill") +
  geom_col(data = p_dat %>% dplyr::filter(plot == "n cells"),
           position = "stack") +
  theme_classic() +
  ggh4x::facet_grid2(plot ~ ., scales = "free_y", independent = "y") +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = event_cols) +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(size = 6))
```

@@

```{r chr7_and_others}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with("_gain"), 
                dplyr::ends_with("_loss")) %>%
  tidyr::pivot_longer(-c("sample", "cell", "cell_group")) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ "WT",
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ","),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <= 25 cells
  dplyr::filter(events != "WT", n >= 25) %>%
  dplyr::mutate(
    events = events %>% 
      gsub("chr", "", .) %>% 
      gsub("_gain", "+", .) %>%
      gsub("_loss", "-", .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, " (", round(prop*100, 2), "%)")
    ))

# plot
p <-
  p_dat %>%
  ggplot(aes(x = events)) +
  geom_text(aes(y = n,  label = label), 
            vjust = 0.5, hjust = 0, angle = 90, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = "bottom")
p + 
  geom_bar(aes(fill = sample))
p + 
  geom_bar(aes(fill = cell_group)) +
  theme(legend.position = "none")
```

@@

```{r cont_tab}
cont <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(chr7 = ifelse(chr7_gain == 1, "chr7+", "chr7_WT"),
                chr5 = ifelse(chr5_gain == 1, "chr5+", "chr5_WT")) 
table(cont$chr7, cont$chr5) %>%
  knitr::kable()
```

@@

  fisher <-
    colnames(sce@colData)[grepl("_gain$|_loss$", colnames(sce@colData))] %>%
    purrr::set_names() %>%
    purrr::map(function(event) {

@@

        # get contingencies
        cont <-
          table(sce@colData[, c("chr7_gain", event)])

         # return relevant stats
        tibble::tibble(
          OR = fisher$estimate, 
          p = fisher$p.value,
          sig = ifelse(fisher$p.value < 0.05, "*", ""),
          n_WT = cont[1, 1],
          n_chr7_gain = cont[2, 1],
          n_event = cont[1, 2],
          n_both = cont[2, 2])

@@

    }) %>%
    dplyr::bind_rows(.id = "event") %>%
    # remove chr7+ (can't do chr7+ vs chr7+)
    dplyr::filter(event != "chr7_gain", n_event > 0) %>%
    dplyr::arrange(p)

@@

}, file = "fisher.rds", rerun = T)

@@

# create glm
# prop_chr7_gain ~ age (+ coverage) (per donor)
for_glm <-
  dplyr::inner_join(
    tibble::as_tibble(sce@colData), 
    metadata, 
    by = c("sample" = "id")) %>%
  dplyr::filter(!is.na(age), !is.na(chr7_gain)) %>%
  dplyr::group_by(sample, age, CTAP, sex, race, ethnicity) %>%
  dplyr::summarise(n_chr7_gain = sum(chr7_gain),
                   n_cells = dplyr::n(),
                   ch7_gain_prop = n_chr7_gain / n_cells)

@@

# all categorical variables
for_glm %>%
  tidyr::pivot_longer(cols = c("CTAP", "sex", "race", "ethnicity")) %>%
  dplyr::group_by(name, value) %>%
  dplyr::mutate(med = median(ch7_gain_prop)) %>%
  ggplot(aes(
    x = tidytext::reorder_within(value, by = med, within = name), 
    y = ch7_gain_prop)) +
  geom_boxplot() +
  geom_jitter(height = 0) +
  facet_wrap(~ name, nrow = 1, scales = "free") +
  tidytext::scale_x_reordered(guide = guide_axis(angle = -45)) +
  theme_classic() +
  labs(x = "")
